---
output:
  html_document: 
    df_print: paged
  pdf_document: default
---
```{r setoptions,echo=FALSE,include=FALSE}
library(caret); library(tidyverse); library(randomForest)
library(knitr); library(xtable)
library(grid); library(gridExtra)
options(xtable.comment = FALSE)
```

```{r computetime,echo=FALSE}
dateSubmit <- format(Sys.time(), "%b %d %Y")
rand <- rnorm(1)
```
# Machine Learning Study: Weightlifting Form Identification
Submitted by Andrew Chang on `r dateSubmit`

## Overview of the Study 
The purpose of this study is to train a model that could identify a specific exercise movement based on kinematic (accelerometer and gyroscopic) measurements and evaluate its predictive ability on a test set of 20 observations. The study determined relevant features extracted from a set of 160 candidate features that were associated with kinematic movements, ultimately arriving at a set of [~50] features as viable predictors for model fitting. In the course of identification, a set of highly correlated predictors were flagged out of the set of viable predictors to better inform model selection decisions regarding bias-reduction/variance-increase trade-offs at a later stage, for either Principal Components Analysis (PCA) or outright elimination. The study then trained three random forest models corresponding to datasets that (1) included, (2) excluded, and (3) transformed the flagged predictors via PCA. The accuracy of three models was then compared and one was selected to identify the 20 exercise movements. 

 
